---
title: Prokaryotic Genome GC Content
author: John Christian Gaby
date: '2020-09-15'
slug: prokaryotic-genome-gc-content
categories: []
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2020-09-15T22:53:36+02:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
draft: true
---

## Introduction

## Genome GC Composition Variation

In a previous blogpost, I examined a prokaryotic genome metadata table from NCBI in order to determine the range of prokaryotic genome sizes. The same data table has the GC% for each prokaryotic genmoe. I will now examine the range of GC contents for the prokaryotes represented in that table.

```{r GCHistogram, message=FALSE}
library(ggplot2)
library(tidyverse)

setwd("/home/chrisgaby/github/My_Website/")

# Read the table into a tibble.
prokaryote.genomes.table <- read_csv(file = "./static/files/prokaryotes.csv")

# Display the minimum, maximum, mean, median, and 1st and 3rd quartiles for the genome sizes.
summary(prokaryote.genomes.table$`GC%`)

```

Now, let's take a look at the distribution of GC%.

```{r}
library(ggplot2)

qplot(data = prokaryote.genomes.table, 
      x = `GC%`, 
      binwidth = 1)

```

The distribution is bimodal. Let's see if that breaks down according to phylum.

```{r}
library(ggplot2)

# Split the Organism Groups column into new columns named according to their respective taxonomy level
prokaryote.genomes.table.split <- separate(data=prokaryote.genomes.table, 
                                           col = `Organism Groups`, 
                                           sep = ";", 
                                           into = c("Domain", "Phylum", "Class"))

phyla.subset.names <- names(which(summary(factor(prokaryote.genomes.table.split$Phylum)) > 90))
prokaryote.genomes.table.split.reduced <- prokaryote.genomes.table.split[prokaryote.genomes.table.split$Phylum == phyla.subset.names[1] | 
                                 prokaryote.genomes.table.split$Phylum == phyla.subset.names[2] | 
                                 prokaryote.genomes.table.split$Phylum == phyla.subset.names[3] | 
                                 prokaryote.genomes.table.split$Phylum == phyla.subset.names[4] | 
                                 prokaryote.genomes.table.split$Phylum == phyla.subset.names[5] | 
                                 prokaryote.genomes.table.split$Phylum == phyla.subset.names[6],]

qplot(data=prokaryote.genomes.table.split.reduced, 
      x = `GC%`, 
      geom = "density", 
      color = Phylum, 
      fill = Phylum, 
      alpha = I(0.2))
```

It looks like some phyla have a bimodal GC% distribution too. Let's break them up by facets.

```{r, fig.height = 8, fig.width = 10, fig.align = "center"}
library(ggplot2)

qplot(data=prokaryote.genomes.table.split.reduced, 
      x = `GC%`, 
      geom = "density", 
      color = Phylum, 
      fill = Phylum, 
      facets = Phylum ~ .)

```
